---
title: "Styling Your Plots"
subtitle: "GVPT399F: Power, Politics, and Data"
format: 
  revealjs:
    slide-number: true
    preview-links: auto
    theme: solarized
    embed-resources: true
execute: 
  echo: true
  fig-align: "center"
---

```{r}
#| echo: false

library(tidyverse)
library(ggthemes)
```

## Data visualization

We will use data visualization to answer the following question:

> Do cars with big engines use more fuel than cars with small engines?

## Add useful titles and labels

```{r}
#| eval: false
#| code-line-numbers: "4,5,6,7,8"

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(colour = class)) + 
  geom_smooth(method = "lm") + 
  labs(
    title = "Engine displacement and highway miles per gallon",
    subtitle = "Values for seven different classes of cars",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  )
```

## Add useful titles and labels

```{r}
#| echo: false

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(colour = class)) + 
  geom_smooth(method = "lm") + 
  labs(
    title = "Engine displacement and highway miles per gallon",
    subtitle = "Values for seven different classes of cars",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  )
```

## Flexible visualization

You can use visual elements to communicate your findings in engaging ways.

```{r}
#| code-line-numbers: "2"

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class == "2seater")) + 
  geom_smooth(method = "lm") + 
  labs(
    title = "Engine displacement and highway miles per gallon",
    subtitle = "Values for seven different classes of cars",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  )
```

## Changing the look of your plots

```{r}
#| code-line-numbers: "2"

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(colour = "pink") + 
  geom_smooth(method = "lm") + 
  labs(
    title = "Engine displacement and highway miles per gallon",
    subtitle = "Values for seven different classes of cars",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  )
```

## Let's clean our graph up

Less is more when it comes to data visualization.

```{r}
#| eval: false
#| code-line-numbers: "4"

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(colour = class)) + 
  geom_smooth(method = "lm") + 
  theme_minimal() + 
  labs(
    title = "Engine displacement and highway miles per gallon",
    subtitle = "Values for seven different classes of cars",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  ) 
```

::: aside
All pre-set `ggplot` themes can be found in the [`ggplot` documentation](https://ggplot2.tidyverse.org/reference/ggtheme.html).
:::

## Let's clean this up

```{r}
#| echo: false

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(colour = class)) + 
  geom_smooth(method = "lm") + 
  theme_minimal() + 
  labs(
    title = "Engine displacement and highway miles per gallon",
    subtitle = "Values for seven different classes of cars",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  )
```

## Creating your own theme

```{r}
#| eval: false
#| code-line-numbers: "4-10"

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(colour = class)) + 
  geom_smooth(method = "lm") + 
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold")
  ) + 
  labs(
    title = "Engine displacement and highway miles per gallon",
    subtitle = "Values for seven different classes of cars",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  ) 
```

::: aside
All theme elements that you can control can be found in the [`ggplot` documentation](https://ggplot2.tidyverse.org/reference/theme.html#arguments). There are a lot!
:::

## Creating your own theme

```{r}
#| echo: false

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(colour = class)) + 
  geom_smooth(method = "lm") + 
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold")
  ) + 
  labs(
    title = "Engine displacement and highway miles per gallon",
    subtitle = "Values for seven different classes of cars",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  )
```

## The before shot

```{r}
#| echo: false

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(colour = class)) + 
  geom_smooth(method = "lm") + 
  labs(
    title = "Engine displacement and highway miles per gallon",
    subtitle = "Values for seven different classes of cars",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  )
```
