---
title: "Plotting Your Data"
subtitle: "GVPT399F: Power, Politics, and Data"
format: 
  revealjs:
    slide-number: true
    preview-links: auto
    theme: solarized
    embed-resources: true
execute: 
  echo: true
  fig-align: "center"
---

```{r}
#| echo: false

library(tidyverse)
```

## Data visualization

We will use data visualization to answer the following question:

> Do cars with big engines use more fuel than cars with small engines?

## Set up your plot

*An empty canvas!*

```{r}
ggplot(data = mpg)
```

## Map your aesthetics

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))
```

## Add in your cars

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point()
```

## Look at the relationship across all cars

```{r}
#| code-line-numbers: "3"

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

## Let's look at groups in the data

-   Can look at more than two interesting elements of our data.

-   You can use visual elements or aesthetics (`aes`) to communicate many dimensions in your data.

-   Let's look at a categorical variable: the class of car (SUV, 2 seater, pick up truck, etc.).

-   *Look for meaningfully defined groups.*

## Let's look at groups in the data

```{r}
#| code-line-numbers: "1"

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, colour = class)) + 
  geom_point()
```

## Look at the relationship within groups

```{r}
#| code-line-numbers: "1,3"

ggplot(data = mpg, mapping = aes(x = displ, y = hwy, colour = class)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

## Aesthetics can be isolated

```{r}
#| code-line-numbers: "2"

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(aes(colour = class)) + 
  geom_smooth(method = "lm")
```
