---
title: "Loading and Exploring Your Data"
subtitle: "GVPT399F: Power, Politics, and Data"
format: 
  revealjs:
    slide-number: true
    preview-links: auto
    theme: solarized
    embed-resources: true
execute: 
  echo: true
  fig-align: "center"
---

```{r}
#| echo: false

library(tidyverse)
library(ggthemes)
```

## Data visualisation

We will use data visualization to answer the following question: 

> Do cars with big engines use more fuel than cars with small engines?

## Load relevant packages 

```{r}
#| eval: false

# Installing new packages
install.packages("ggthemes")

# Load the relevant packages
library(tidyverse)
library(ggthemes)
```

## Loading in relevant data

```{r}
#| eval: false

# Load the data
mpg
```

```{r}
#| echo: false

mpg |> 
  head() |> 
  select(1:5) |> 
  knitr::kable()
```

## EXERCISE {background-color="#B56576"}

<br>

Learn more about this data set by typing `?mpg` into your console.

::: aside
To get help on any function or data set use: `?function`
:::

## The `mpg` data set

```{r}
glimpse(mpg)
```

## The `mpg` data set

A couple of useful variables:

-   `displ`: engine displacement, in litres

-   `hwy`: highway miles per gallon

## EXERCISE {background-color="#B56576"}

1.  How many rows are in `mpg`? How many columns?

```{r}
#| eval: false

nrow(mpg)
ncol(mpg)
```

<br>

2.  What does the `drv` variable describe?

```{r}
#| eval: false

?mpg
```

