---
title: "Filtering Your Data"
subtitle: "GVPT399F: Power, Politics, and Data"
format: 
  revealjs:
    slide-number: true
    preview-links: auto
    theme: solarized
    embed-resources: true
execute: 
  echo: true
---

## `dplyr` basics

-   First argument is always a data object (for example, a dataframe).

-   Subsequent arguments typically describe which columns to operate on, using the variable names (without quotes).

-   Output is always a new data object.

## Filter rows with `filter()`

```{r}
#| echo: false

library(tidyverse)
library(gapminder)

```

```{r}
filter(gapminder, country == "Australia", year > 2000)
```

## Filter rows with `filter()`

```{r}
filter(gapminder, continent %in% c("Asia", "Oceania"))
```

## Filter rows with `filter()`

```{r}
filter(gapminder, pop > 500000 & pop < 1000000)
```

## Filter rows with `filter()`

```{r}
filter(gapminder, pop > 500000 | pop < 1000000)
```

## Handy operations

`==` is *equal to*

<br>

`!=` is *not equal to*

<br>

`>=` is *greater than or equal to*

<br>

`<=` is *less than or equal to*

## Handy operations

<br>

`|` is *OR*

<br>

`&` is *AND*

<br>

`%in%` is *in*
