---
title: "From Surveys to Populations"
subtitle: "GVPT399F: Power, Politics, and Data"
format: 
  revealjs:
    slide-number: true
    preview-links: auto
    theme: solarized
    embed-resources: true
execute: 
  echo: true
  warning: false
  fig-align: "center"
bibliography: references.bib
---

```{r}
#| echo: false 

library(tidyverse)
```

## Surveys

-   Populations are very difficult to collect data on

    -   Even the census misses people!

-   Happily, we can use surveys of a sample of our population to learn things about our population

-   However, our ability to do this is conditional on how good our sample is

-   What do I mean by "good"?

## The 2024 US Presidential Election

-   Elections are preceded by a flood of surveys

## Surveys

-   Surveys are conducted on a subset (**sample**) of the population of interest

-   Our population of interest: individuals who voted in the 2024 US Presidential Election

## A good sample

-   A good sample is a **representative one**

-   How closely does our sample reflect our population

## Parallel worlds

-   Remember back to last session on experiments

-   In an ideal world, we would be able to create two parallel worlds (one with the treatment, one held as our control)

    -   One version of the election booth run without monitors (the control)

    -   One version with monitors (the treatment)

-   These worlds are perfectly identical to each other prior to treatment

-   We cannot do this :(

## The next best thing

-   Our next best option is to create two groups that were as identical to one another as possible prior to treatment

-   If they are (almost) identical, differences between their group-wide outcomes can be attributed to the treatment

-   One good way of getting two (almost) identical groups is to assign individuals to those groups **randomly**

    -   Think back to our 1,000 hypothetical people!

## Randomization

-   Randomization continues to pop its chaotic head up

-   We can use it to create a sample that is (almost) identical to our population, on average

-   Drawing randomly from our population increases our chances of ending up with a sample that reflects that population

-   This would be referred to as a **representative sample**

## Random sampling

-   All individuals in the population need to have an equal chance of being selected for the sample

    -   If this holds, you have a **pure random sample**

-   This is really hard to do!

    -   How likely were you to answer the pollster's unknown number, calling you in the middle of the day?

    -   Even if you did answer, how likely were you to answer all their questions?

## To illustrate

Countries' GDP in 2022:

```{r}
#| echo: false

gdp_df <- wbstats::wb_data("NY.GDP.MKTP.CD", 
                           return_wide = F,
                           start_date = 2022,
                           end_date = 2022) |> 
  mutate(sample_value = if_else(value < 31514830422 & rbinom(1, 1, 0.5) == 1,
                                NA,
                                value))

ggplot(gdp_df, aes(x = value / 1e9)) + 
  geom_histogram() + 
  theme_minimal() + 
  labs(x = "GDP (USD billions)",
       y = "Count") +
  scale_x_continuous(labels = scales::dollar)
```

## Countries' GDP

I want to estimate the average GDP across all countries in 2022.

-   I send out a survey to all countries' Departments of Statistics and ask for their GDP figures for 2022.

-   I get 60 responses:

```{r}

sample_df <- gdp_df |> 
  drop_na(sample_value) |> 
  sample_n(size = 60) |> 
  transmute(country, gdp = sample_value)

sample_df
```

## Countries' GDP

I now calculate the average of these responses, which I find to be:

```{r}

sample_df |> 
  summarise(avg_gdp = scales::dollar(mean(gdp, na.rm = T)))
```

Now, imagine that we knew definitively that it was `r scales::dollar(mean(gdp_df$value))`. Why such a large difference?

## Non-response bias

Poorer countries are far less likely to be able or willing to provide these economic data to academics or international organizations.

-   They tend to be underrepresented in a lot of data

My sample was biased against poorer countries.

-   They were not equally likely to respond to my request for data as rich countries

## Large numbers

-   Randomization isn't enough: we also need to draw a sufficiently large sample from our population

    -   One person pulled randomly from the class isn't going to be very reflective of the class!
