---
title: "Data Transformation"
subtitle: "GVPT399F: Power, Politics, and Data"
format: 
  revealjs:
    slide-number: true
    preview-links: auto
    theme: solarized
    embed-resources: true
execute: 
  echo: true
---

## Learning objectives

1.  Learn basic operations in R

2.  Be introduced to `dplyr`

3.  Clean up and transform your data

## R objects

Create new objects with `<-`

```{r}
x <- 3 * 4

x
```

<br>

```{r}
x <- 3 * 10

x
```

::: aside
Some people use `=` instead of `<-`. I strongly recommend against this. It makes your script difficult to read, and it can lead to syntax errors.
:::

## R functions

Many functions come with R straight out of the box:

```{r}
seq(1, 10)
```

<br>

You can create objects using functions:

```{r}
x <- seq(1, 10)

x
```

## Gapminder

First, you need to install the `gapminder` package:

```{r}
#| eval: false

install.packages("gapminder")
```

Then access the `gapminder` data set:

```{r}
library(tidyverse)
library(gapminder)

head(gapminder)
```

## Data types

In `gapminder`:

-   `fctr` stands for factors, which R uses to represent categorical variables with fixed possible values.

-   `int` stands for integer.

-   `dbl` stands for doubles (or real numbers).

## Data types

Other types:

-   `chr` stands for character vectors, or strings.

-   `dttm` stands for date-times (a date + a time).

-   `lgl` stands for logical, vectors that contain only `TRUE` or `FALSE`.[^1]

[^1]: `TRUE` can be shortened to `T` and `FALSE` can be shortened to `F`.

## Introducing `dplyr`

*Help you with most of your data transformation needs.*

Five basic functions:

-   `filter()`

-   `arrange()`

-   `select()`

-   `mutate()`

-   `summarise()`
