---
title: "The pipe"
subtitle: "GVPT399F: Power, Politics, and Data"
format: 
  revealjs:
    slide-number: true
    preview-links: auto
    theme: solarized
    embed-resources: true
execute: 
  echo: true
---

```{r}
#| echo: false

library(tidyverse)
library(gapminder)
```

## Combine multiple operations with the pipe

That got messy!

-   We had a lot of different objects representing intermediate steps in our calculations.

-   We never need those objects again. Can we avoid creating them?

Let's introduce perhaps the defining feature of the tidyverse: **the pipe**.

## Combine multiple operations with the pipe

Read the pipe as:

Take this \|\> (*and then...*) <br>     do this \|\> (*and then...*) <br>     do this

```{r}
gapminder |> 
  group_by(continent) |> 
  summarise(avg_pop = mean(pop), avg_gdp_per_cap = mean(gdpPercap)) |> 
  arrange(avg_gdp_per_cap)
```

## Combine multiple operations with the pipe

```{r}
gapminder |> 
  group_by(continent) |> 
  summarise(avg_pop = mean(pop), avg_gdp_per_cap = mean(gdpPercap)) |> 
  ggplot(aes(x = continent, y = avg_gdp_per_cap)) + 
  geom_col() + 
  theme_minimal()
```

## A note on the pipe

::::: columns
::: {.column width="50%"}
Base pipe:

-   `|>`

-   Can be used without loading any packages

-   Relatively new: introduced in 2021
:::

::: {.column width="50%"}
Tidyverse pipe:

-   `%>%`

-   Must load `dplyr` or `magrittr` to use
:::
:::::

## Summary

This session you have:

1.  Learnt R basic syntax

2.  Learnt how to transform your data

3.  Written concise code that is easy to follow
